Scaling rates of true polar wander in convecting planets and moons

### 2017-05-01 tpw_rate.tex
More edits from Bruce.

\begin{abstract}
Mass redistribution in the convecting mantle of a planet causes perturbations in its moment of inertia tensor. 
Conservation of angular momentum dictates that these perturbations change the direction of the rotation vector of the planet, a process known as true polar wander (TPW). 
Although the existence of TPW on Earth is firmly established, its rate and magnitude over geologic time scales remain controversial. 
Here we present scaling analyses and numerical simulations of TPW due to mantle convection over a range of parameter space relevant to planetary interiors. 
For simple rotating convection, we identify a set of dimensionless parameters that characterize true polar wander. We use these parameters to define  timescales for the growth of moment of inertia perturbations due to convection and for their relaxation due to true polar wander. 
These timescales, as well as the relative sizes of convective anomalies, control the rate and magnitude of TPW.
This analysis also clarifies the nature of so called ``inertial interchange'' TPW events, and relates them to a  broader class of events that enable large and often rapid TPW. We expect these events to have been more frequent in Earth’s past.
\end{abstract}

@@

We are in a position to address the rates of true polar wander for a given convective moment $\mathbf{E}$.
A considerable simplification occurs if we neglect secular changes in the rotation rate, and just consider changes in direction of the pole ($d \Omega^2 / dt = 2 {\Omega_i} \dot{ \Omega}_i = 0$)
Substituting Equation~\eqref{eq:rotational_deformation} into Equation~\eqref{eq:disequilibrium} we find
\begin{equation}
\frac{k_f a^5 T_1 \Omega^2}{3G}\mitbf{\Omega} \times \dot{\mitbf{\Omega} } = \mitbf{\Omega} \times \left( \mathbf{E} \cdot \mitbf{\Omega} \right).
\label{eq:liouville}
\end{equation}
Introducing a unit vector $\mitbf{\omega} = \mitbf{\Omega}/\|\mitbf{\Omega} \|$ and using Equation~\eqref{eq:fluid_love} for $k_f$,  we may solve this equation for $\dot{\mitbf{\omega}}$:
\begin{equation}
 \dot{\mitbf{\omega}}  = \frac{1}{(C-A)T_1} \left[ \mathbf{E} \cdot \mitbf{\omega} - \left( \mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega}  \right) \mitbf{\omega} \right].
\label{eq:liouville_ode}
\end{equation}
In order to evaluate $\dot{\mitbf{\omega}}$ we require knowledge of both $\mathbf{E}$ and $\mitbf{\omega}$ in the geographic frame (i.e. the frame used to derive Equation (13)). The same frame is also used to measure TPW, but the expressions on the right-hand-side of Equation (13) are more naturally evaluated in the reference frame of the principal axes of $\mathbf{E}$.
We can avoid the need to reconcile these two frames by confining our attention to the rate of polar wander (We denoted  the rate of TPW by 
 $\dot{W} = \vert \dot{\mitbf{\omega}} \vert$) .
Evaluating the scalar product $\dot{W}^2 = \dot{\mitbf{\omega}} \cdot \dot{\mitbf{\omega}}$ gives
\begin{equation}
\dot{\Theta}^2 = \dot{\mitbf{\omega}}^2 = \frac{1}{(C-A)^2 T_1^2} \left[ (\mathbf{E}\cdot\mitbf{\omega})^2 - (\mitbf{\omega}\cdot\mathbf{E}\cdot\mitbf{\omega})^2 \right].
\label{eq:polar_wander_rate_tensor}
\end{equation}
The scalar terms on the right-hand-side of Equation (14) are now independent of the choice of coordinate system as long as we are consistent.
Therefore we can enter the coordinate system of the convective moment of inertia $\mathbf{E}$ 
with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$. In this case the orientation of 
$\mitbf{\omega}$ is defined in the $\mathbf{E}$-frame using colatitude $\theta$ and longitude $\phi$ (see Figure~\ref{fig:reference_frames}).
Plugging this description of $\mitbf{\omega}$ into Equation~\eqref{eq:polar_wander_rate_tensor},
and after some tedious algebra, we find

\begin{equation}
\begin{aligned}
\dot{W}^2 = 
&\frac{1}{4(C-A)^2 T_1^2}\sin^2{2\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] +\\
&\frac{1}{4(C-A)^2 T_1^2}\sin^4{\theta} \sin^2{2\phi}\left( \lambda_2 - \lambda_1 \right)^2. \\
\label{eq:polar_wander_rate}
\end{aligned}
\end{equation}
This equation is a version of what has been called the ``Milankovitch theorem'' \citep{munk1960rotation}. Once the orientation of $\mitbf{\omega}$ is specified in the E-frame by angles $\theta$ and $\phi$, we can evaluate the rate of TPW in terms of the principal moments of $\mathbf{E}$.
Two special cases of this equation are worth  noting. First, if $\mitbf{\omega}$  coincides with  the normal to the octahedral plane
(i.e. the plane defined by a normal with direction cosines all equal to $1/\sqrt{3}$, or $\phi=45^\circ, \theta\approx 55^\circ$, cf. \citet{fung1965foundations}), 
then it can be expressed in terms of the second invariant ($E_{\mathrm{II}}$) of the moment of inertia deviator 
($E_{ij} - 1/3 E_{kk} \delta_{ij}$):
\begin{equation}
\begin{aligned}
\dot{W}^2
&= \frac{1}{9(C-A)^2 T_1^2}\left[( \lambda_3 - \lambda_1)^2 + (\lambda_3 - \lambda_2)^2 + (\lambda_2 - \lambda_1)^2 \right]\\
&= \frac{E_{\mathrm{II}}}{9(C-A)^2 T_1^2}. \\
\end{aligned}
\end{equation}
The second invariant of the stress deviator is commonly used in the theories of plasticity
to characterize the state of shear stress.
The second invariant of the moment of inertia deviator can be viewed as a ``rotational stress" that drives a TPW response.

@@

Returning to the three-dimensional problem we expect $\Lambda_{ij}$ to scale as
\begin{equation}
\Lambda_{ij} \sim \frac{\Gamma}{\mathrm{Ra} }.
\label{eq:lambda_estimate}
\end{equation}
Using this estimate in Equation (33) gives
\begin{equation}
\frac{d W}{dt^\prime} = \frac{1}{m} \sin 2 \theta
\end{equation}
Remarkably, $Ra$ and $\Gamma$ have completely dropped from the expression for the rate of TPW.  The time constant $\tau_R$ goes down at high Rayleigh numbers. At the same time, the  coherence in the temperature structure goes down, reducing the amount of power in the degree-two part of the field responsible for driving TPW.
That these two effects cancel is something of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
Scalings for lengthscales of convection in fluids with temperature dependent viscosity \citep[e.g.][]{solomatov1995scaling} or pseudoplastic rheology \citep[e.g.][]{korenaga2010scaling} have different functional dependencies on $\mathrm{Ra}$ or additional nondimensional parameters.
However, a common feature in most scalings is that typical lengthscales are still some power-law of Rayleigh number $d \propto \mathrm{Ra}^{-\beta}$.
With this form, our scaling for TPW rate has the following dependence on $\mathrm{Ra}$:
\begin{equation}
\frac{d\Theta}{dt^\prime} \sim \frac{\mathrm{Ra}^{1-3\beta}}{m} \sin{2 \theta}.
\label{eq:simplest_milankovitch_defer_scaling}
\end{equation}
In general, $\beta$ is some small number between one-fourth and one-third, so we expect that more complicated estimates for 
$d(\mathrm{Ra})$ will still result in a weak dependence 
on the Rayleigh number.

@@

Given a random perturbation to $\mathbf{E}$, we would like to give a bound on the size of changes in the orientation of the principle axes. Let $\delta$ be the size of the perturbation to the convective moment of inertia tensor after some time interval $\Delta t$, and let $\lambda_3 \ge \lambda_2 \ge \lambda_1$ be the eigenvalues of that tensor.  The corresponding rotation of the principal axes is defined by the angle $\xi$. A bound on $\xi$ is given by \citep{davis1970rotation}
\begin{equation}
\vert \sin(2 \xi) \vert \le \frac{ 2 \vert \delta \vert}{ \displaystyle \min_{i \neq j} \vert \lambda_i - \lambda_j \vert }.
\label{eq:kahan}
\end{equation} 
Small changes in orientation (i.e. small $\xi$) occur when the difference in eigenvalues is large. Conversely, large changes in orientation are possible when differences between the eigenvalues are small. 
This dependence is illustrated schematically in Figure~\ref{fig:perturb}.  Unstable conditions prevail when the maximum and intermediate moments of inertia are nearly equal, leading to large TPW events \citep{evans1998true}. (\citet{tsai2007theoretical} argue that large TPW events take a long time because the rate of TPW is small when $\Lambda_{31}$ is small or when $\theta = 0$ or $\pi/2$ (see Equation (33). If the maximum and intermediate axes are interchanged when the misalignment angle is small $\theta \approx 0$, then the new definition of the maximum axis after the interchange ensures that $\theta \approx \pi/2$. In this case the rate of TPW may be small before and after the interchange).

@@

The angle $\xi$ would correspond to a change in $\theta$ if relaxation was not allowed to operate. While this approximation is not realistic, it does suggest that larger perturbations in $\theta$ are possible when  $(\lambda_i-\lambda_j)$ small. Scaling results and numerical simulations from the previous section show that the difference in eigenvalues scales with $\sim \mathrm{Ra}^{-1}$.
Therefore, the characteristic gap between the eigenvalues of the convective moment becomes smaller as the Rayleigh number increases.
Additionally, the timescale of fluctuations in these values goes down because the density anomalies are moving faster.
Overall, this makes the principal axes of high Rayleigh number systems much less stable, consistent with the result of \citet{richards1999polar}.

@@

Figure~\ref{fig:misfit} shows a representative time series from the convection calculations. As before, we compute the convective moment of inertia in the geographic frame and evaluate the eigenvalues of $\mathbf{E}$ to determine the eigengap. We also evolve the orientation of the rotation axis by integrating  Equation~\eqref{eq:liouville_ode} in time.
Since it is a 2D model, the spin axis is described by a single angle. The misalignment angle $\theta$ also lies in the plane of the 2D model.
When the eigengap gets small,  the corresponding misalignment angle increases and the rate of polar wander becomes much faster. This behavior is quantitatively consistent with the predictions from Equation (33). We draw special attention to two events in the time series. At $\sim$0.3 Gyr the eigengap dips and the misalignment angle increases to $\theta \sim 20^{\circ}$. There is a large change in the orientation of the spin axis, even though
there is technically no interchange of the maximum and intermediate axes. A conventional IITPW event occurs at $\sim 1$ Gyr;  the eigengap goes to zero, interchanging the maximum and intermediate axes. The misfit angle goes to approximately $90^\circ$. During this time the orientation of the spin axis changes by roughly 90$^{\circ}$.

Large TPW events are associated with inertial interchanges, but a broader class of events are evident
when the principal moments are close to each other. We argue that large TPW events do not (necessarily) imply an inertial interchange. Conversely, we do not expect all large TPW events to be associated with a 90$^{\circ}$ change in the spin. Our scaling results suggest that large TPW events are more likely to occur in the early Earth when the planet is convecting more vigorously and the characteristic gap between principal moments is small.

\citet{tsai2007theoretical} suggested that interchanges in the principal moments are unlikely to produce large
TPW events because the rate of TPW is low. This implies that a large 90$^{\circ}$ change in the orientation of the rotation axis occurs over a long time.  While
this is true, the numerical convection calculation reveals a more complex behavior. 
First, the principal moments can interchange, or almost interchange, quite rapidly. In these cases  the perturbations in the convective moments can become large soon after an interchange. Second, the
 the location of the principal axes do not stay put during a IITPW event. A persistent drift of the principal axes can can lead to extended polar wandering, yielding angular changes greater than 90$^\circ$.

@@

Indeed, we can identify two endmember behaviors of Equation~\eqref{eq:simplest_milankovitch_defer_scaling}.
When convection is not sufficiently chaotic to create a large $\theta$ we are in the regime where the planet's rotation axis closely tracks that of the convective moment.
This is the regime considered in \citet{steinberger1997changes}, \citet{roberts2007cause}, and \citet{zhong2007supercontinent}, and can be considered the ``slow TPW'' regime.
When convection is more chaotic, however, there may be large excursions in $\theta$, 
which are driven by large changes in the orientation of the principal axes when $(\lambda_i - \lambda_j)$ is small.  
In this case a dramatic increase in the rate of TPW is possible. We refer to this as the `` fast TPW’’. Inertial interchanges occur when the maximum and intermediate axis are swapped and this can lead to a large change in 
 $\theta$ \citep{kirschvink1997evidence}.  Increasing $\theta$ to $\pi/2$ does not necessarily require fast TPW, but our numerical simulations show that this is often the case. Moreover, the inertial interchange is only  a special case of the large $\theta$, ``fast TPW'' regime. We expect large $\theta$ and ``fast TPW’’ to be more prevalent in the early Earth, when the mantle was presumably hotter and less viscous (i.e. higher $Ra$). 
More vigorous convection increases the rate of change of density anomalies  and  makes
$\theta(t)$ less stable. Large (and potentially fast) TPW events should have been more frequent.

For Cenozoic Earth we can substitute direct estimates of the important parameters into Equation~\eqref{eq:simple_milankovitch}.
Typical values for the time constant $\tau_R$ are of order $30 \, \mathrm{kyr}$ \citep{ricard1993polar}.
Estimates of the present day non-hydrostatic moment of inertia (due to mantle density anomalies, corresponding to $\Lambda_{31}I_0$ in the preceding scaling)
are in the neighborhood of $10^{-5} I_0$, while the hydrostatic moment of inertia (corresponding to $C-A$) is $3 \times 10^{-3} I_0$ \citep{chambat2001mean}.
A key question is whether convection is sufficiently chaotic to enter the large $\theta$ regime. 
\citet{richards1997explanation} have argued that the convective planform of Earth has been stable for the last
few hundred million years. On the other hand, we know that there have been large reorganizations of that planform 
during Earth history, so this recent geologic stability may not hold in general \citep{evans2003true}.
Our numerical simulations show that large values for $\theta$ are possible in a vigorously convecting mantle.
Allowing for such a large mismatch angle ($\theta = 45^\circ$) we may estimate the maximum polar wander rate
\begin{equation}
\max ( \dot{\Theta} ) = \frac{(\lambda_3-\lambda_1)}{(C-A)}\frac{1}{T_1} \sim 6^\circ / \mathrm{Myr},
\end{equation}
which corresponds to about 66 cm/yr 90$^\circ$ from the TPW axis.
This value is similar to the rates discussed by \citet{cambiotti2011new} for the past 100 Myr, though our rate is roughly a factor of four larger because we allow for the
possibility of a larger mismatch angle. Such large angles may be necessary to explain the range of rates suggested by some interpretations of paleomagnetic data \citep{mitchell2011sutton}.
The bulk viscosity of Earth's mantle is uncertain by up to a factor of ten \citep{mitrovica2004new},
which results in a corresponding uncertainty for the relaxation time $T_1$ and the maximum polar wander rate.

>>>>

\begin{abstract}
Mass redistribution in the convecting mantle of a planet causes perturbations in its moment of inertia tensor. 
Conservation of angular momentum dictates that these perturbations change the direction of the rotation vector of the planet, a process known as true polar wander (TPW). 
Although the existence of TPW on Earth is firmly established, its rate and magnitude over geologic time scales remain controversial. 
Here we present scaling analyses and numerical simulations of TPW due to mantle convection over a range of parameter space relevant to planetary interiors. 
For simple rotating convection, we identify a set of dimensionless parameters that fully characterize true polar wander. We use these parameters to define  timescales for the growth of moment of inertia perturbations due to convection and for their relaxation due to true polar wander. 
These timescales, as well as the relative sizes of convective anomalies, control the rate and magnitude of TPW.
This analysis also clarifies the nature of so called ``inertial interchange'' TPW events, and relates them to a  broader class of events that enable large and often rapid TPW. We expect these events to have been more frequent in Earth’s past.
\end{abstract}

@@

We are in a position to address the rates of true polar wander for a given convective moment $\mathbf{E}$.
A considerable simplification occurs if we neglect secular changes in the rotation rate, and just consider changes in direction of the pole ($d \Omega^2 / dt = 2 {\Omega_i} \dot{ \Omega}_i = 0$)
Substituting Equation~\eqref{eq:rotational_deformation} into Equation~\eqref{eq:disequilibrium} we find
\begin{equation}
\frac{k_f a^5 T_1 \Omega^2}{3G}\mitbf{\Omega} \times \dot{\mitbf{\Omega} } = \mitbf{\Omega} \times \left( \mathbf{E} \cdot \mitbf{\Omega} \right).
\label{eq:liouville}
\end{equation}
Introducing a unit vector $\mitbf{\omega} = \mitbf{\Omega}/\|\mitbf{\Omega} \|$ and using Equation~\eqref{eq:fluid_love} for $k_f$,  we may solve this equation for $\dot{\mitbf{\omega}}$:
\begin{equation}
 \dot{\mitbf{\omega}}  = \frac{1}{(C-A)T_1} \left[ \mathbf{E} \cdot \mitbf{\omega} - \left( \mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega}  \right) \mitbf{\omega} \right].
\label{eq:liouville_ode}
\end{equation}
In order to evaluate $\dot{\mitbf{\omega}}$ we require knowledge of both $\mathbf{E}$ and $\mitbf{\omega}$ in the geographic frame (i.e. the frame used to derive Equation (13)). The same frame is also used to measure TPW, but the expressions on the right-hand-side of Equation (13) are more naturally evaluated in the reference frame of the principal axes of $\mathbf{E}$.
We can avoid the need to reconcile these two frames by confining our attention to the rate of TPW because scalar quantities are invariant under coordinate rotations. (We denoted  the rate of wander by 
 $\dot{W} = \vert \dot{\mitbf{\omega}} \vert$) .
Evaluating the scalar product $\dot{W}^2 = \dot{\mitbf{\omega}} \cdot \dot{\mitbf{\omega}}$ gives
\begin{equation}
\dot{W}^2 = \dot{\mitbf{\omega}}^2 = \frac{1}{(C-A)^2 T_1^2} \left[ (\mathbf{E}\cdot\mitbf{\omega})^2 - (\mitbf{\omega}\cdot\mathbf{E}\cdot\mitbf{\omega})^2 \right].
\label{eq:polar_wander_rate_tensor}
\end{equation}
The scalar terms on the right-hand-side of Equation (14) can now be evaluated in either the geographic or the $E$ frames.
Therefore we can enter the coordinate system of the convective moment of inertia $\mathbf{E}$ 
with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$. In this case the orientation of 
$\mitbf{\omega}$ is defined in the $\mathbf{E}$-frame using colatitude $\theta$ and longitude $\phi$ (see Figure~\ref{fig:reference_frames}).
Plugging this description of $\mitbf{\omega}$ into Equation~\eqref{eq:polar_wander_rate_tensor},
and after some tedious algebra, we find

\begin{equation}
\begin{aligned}
\dot{W}^2 = 
&\frac{1}{4(C-A)^2 T_1^2}\sin^2{2\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] +\\
&\frac{1}{4(C-A)^2 T_1^2}\sin^4{\theta} \sin^2{2\phi}\left( \lambda_2 - \lambda_1 \right)^2. \\
\label{eq:polar_wander_rate}
\end{aligned}
\end{equation}
This equation is a version of what has been called the ``Milankovitch theorem'' \citep{munk1960rotation}. Once the orientation of $\mitbf{\omega}$ is specified in the E-frame by angles $\theta$ and $\phi$, we can evaluate the rate of TPW in terms of the principal moments of $\mathbf{E}$.
Two special cases of this equation are identified by noting a similarity in the form of Equation (13) to the shear stress on a plane in classical elastostatics. First, if $\mitbf{\omega}$  coincides with  the normal to the octahedral plane
(i.e. the plane defined by a normal with direction cosines all equal to $1/\sqrt{3}$, or $\phi=45^\circ, \theta\approx 55^\circ$, cf. \citet{fung1965foundations}), 
then it can be expressed in terms of the second invariant ($E_{\mathrm{II}}$) of the moment of inertia deviator 
($E_{ij} - 1/3 E_{kk} \delta_{ij}$):
\begin{equation}
\begin{aligned}
\dot{W}^2
&= \frac{1}{9(C-A)^2 T_1^2}\left[( \lambda_3 - \lambda_1)^2 + (\lambda_3 - \lambda_2)^2 + (\lambda_2 - \lambda_1)^2 \right]\\
&= \frac{E_{\mathrm{II}}}{9(C-A)^2 T_1^2}. \\
\end{aligned}
\end{equation}
The second invariant of the stress deviator is commonly used in the theories of plasticity
to characterize the state of shear stress.
The second invariant of the moment of inertia deviator can be viewed as a ``rotational stress" that drives a TPW response.

@@

Returning to the three-dimensional problem we expect $\Lambda_{ij}$ to scale as
\begin{equation}
\Lambda_{ij} \sim \frac{\Gamma}{\mathrm{Ra} }.
\label{eq:lambda_estimate}
\end{equation}
Using this estimate in Equation (33) gives
\begin{equation}
\frac{d W}{dt^\prime} = \frac{1}{m} \sin 2 \theta
\end{equation}
Remarkably, $Ra$ and $\Gamma$ have completely dropped from the expression for the rate of TPW.  The time constant $\tau_R$ goes down at high Rayleigh numbers. At the same time, the  coherence in the temperature structure goes down, reducing the amount of power in the degree-two part of the field responsible for driving TPW.
That these two effects cancel is something of a coincidence due to the simple estimate of the smallest lengthscales of the problem.
Scalings for lengthscales of convection in fluids with temperature dependent viscosity \citep[e.g.][]{solomatov1995scaling} or pseudoplastic rheology \citep[e.g.][]{korenaga2010scaling} have different functional dependencies on $\mathrm{Ra}$ or additional nondimensional parameters.
However, a common feature in most scalings is that typical lengthscales are still some power-law of Rayleigh number $d \propto \mathrm{Ra}^{-\beta}$.
With this form, our scaling for TPW rate has the following dependence on $\mathrm{Ra}$:
\begin{equation}
\frac{d W}{dt^\prime} \sim \frac{\mathrm{Ra}^{1-3\beta}}{m} \sin{2 \theta}.
\label{eq:simplest_milankovitch_defer_scaling}
\end{equation}
In general, $\beta$ is some small number between one-fourth and one-third, so we expect that more complicated estimates for 
$d(\mathrm{Ra})$ will still result in a weak dependence 
on the Rayleigh number.

@@

Given a random perturbation to $\mathbf{E}$, we would like to give a bound on the size of changes in the orientation of the principle axes. Let $\delta$ be the size of the perturbation to the convective moment of inertia tensor after some time interval $\Delta t$, and let $\lambda_3 \ge \lambda_2 \ge \lambda_1$ be the eigenvalues of that tensor.  The corresponding rotation of the principal axes is defined by the angle $\xi$. A bound on $\xi$ is given by \citep{davis1970rotation}
\begin{equation}
\vert \sin(2 \xi) \vert \le \frac{ 2 \vert \delta \vert}{ \displaystyle \min_{i \neq j} \vert \lambda_i - \lambda_j \vert }.
\label{eq:kahan}
\end{equation} 
Small changes in orientation (i.e. small $\xi$) occur when the difference in eigenvalues is large. Conversely, large changes in orientation are possible when differences between the eigenvalues are small. 
This dependence is illustrated schematically in Figure~\ref{fig:perturb}.  Unstable conditions prevail when the maximum and intermediate moments of inertia are nearly equal, leading to large TPW events \citep{evans1998true}. On the hand hand, \citet{tsai2007theoretical} argue that large TPW events take a long time because the rate of TPW is small when $\Lambda_{31}$ is small or when $\theta = 0$ or $\pi/2$ (see Equation (33)). If the maximum and intermediate axes are interchanged when the misalignment angle is small $\theta \approx 0$, then the new definition of the maximum axis after the interchange ensures that $\theta \approx \pi/2$. In this case the rate of TPW may be small before and after the interchange.

@@

The angle $\xi$ would correspond to a change in $\theta$ if relaxation was not allowed to operate. While this approximation is not realistic, it does suggest that larger perturbations in $\theta$ are possible when  $(\lambda_i-\lambda_j)$ small. Scaling results and numerical simulations from the previous section show that the difference in eigenvalues scales with $\sim \mathrm{Ra}^{-1}$.
Therefore, the characteristic gap between the eigenvalues of the convective moment becomes smaller as the Rayleigh number increases.
Additionally, the timescale of fluctuations in these values goes down because the density anomalies are moving faster.
Overall, this makes the principal axes of high Rayleigh number systems much less stable, consistent with the result of \citet{richards1999polar}, based on mantle convection calculations.

@@

Figure~\ref{fig:misfit} shows a representative time series from the convection calculations. As before, we compute the convective moment of inertia in the geographic frame and evaluate the eigenvalues of $\mathbf{E}$ to determine the eigengap. We also evolve the orientation of the rotation axis by integrating  Equation~\eqref{eq:liouville_ode} in time.
Since it is a 2D model, the spin axis is described by a single angle. The misalignment angle $\theta$ also lies in the plane of the 2D model.
When the eigengap gets small,  the corresponding misalignment angle increases and the rate of polar wander becomes much faster. This behavior is consistent with the predictions from Equation (33). We draw special attention to two events in the time series. At $\sim$0.3 Gyr the eigengap dips and the misalignment angle increases to $\theta \sim 20^{\circ}$. There is a large change in the orientation of the spin axis, even though
there is technically no interchange of the maximum and intermediate axes. A conventional IITPW event occurs at $\sim 1$ Gyr;  the eigengap goes to zero, interchanging the maximum and intermediate axes. The misfit angle goes to approximately $90^\circ$. During this time the orientation of the spin axis changes by roughly 90$^{\circ}$.

Large TPW events are associated with inertial interchanges, but a broader class of events are evident
when the principal moments are close to each other. We argue that large TPW events do not (necessarily) imply an inertial interchange. Conversely, we do not expect all large TPW events to be associated with a 90$^{\circ}$ change in the spin axis. Our scaling results suggest that large TPW events are more likely to occur in the early Earth when the planet is convecting more vigorously and the characteristic gap between principal moments is small.

\citet{tsai2007theoretical} suggested that large TPW events do not always accompany an interchange of the principal moments because the rate of TPW is low. A large 90$^{\circ}$ change in the orientation of the rotation axis can only occur over a long time.  While
this is true for idealized cases, the numerical convection calculation reveals a more complex behavior. 
First, the principal moments can interchange, or almost interchange, quite rapidly. In these cases  the perturbations in the convective moments can become large soon after an interchange. Second, the
 the location of the principal axes do not stay put during a IITPW event. A persistent drift of the principal axes can can lead to extended polar wandering, yielding angular changes greater than 90$^\circ$.

@@

Indeed, we can identify two endmember behaviors of Equation~\eqref{eq:simplest_milankovitch_defer_scaling}.
When convection is not sufficiently chaotic to create a large $\theta$ we are in the regime where the planet's rotation axis closely tracks that of the convective moment.
This is the regime considered in \citet{steinberger1997changes}, \citet{roberts2007cause}, and \citet{zhong2007supercontinent}, and can be considered the ``slow TPW'' regime.
When convection is more chaotic, however, there may be large excursions in $\theta$, 
which are driven by large changes in the orientation of the principal axes when $(\lambda_i - \lambda_j)$ is small.  
In this case a dramatic increase in the rate of TPW is possible. We refer to this as the `` fast TPW’’. Inertial interchanges occur when the maximum and intermediate axis are swapped and this can lead to a large change in 
 $\theta$ \citep{kirschvink1997evidence}.  Increasing $\theta$ to $\pi/2$ does not necessarily require fast TPW, but our numerical simulations show that rapid motion often occurs because $\theta$ does not remain close to $\pi/2$.  Moreover, the inertial interchange is only  a special case of the large $\theta$, ``fast TPW'' regime. We expect large $\theta$ and ``fast TPW’’ to be more prevalent in the early Earth, when the mantle was presumably hotter and less viscous (i.e. higher $Ra$). 
More vigorous convection increases the rate of change of density anomalies  and  makes
$\theta(t)$ less stable. Large (and potentially fast) TPW events should have been more frequent.

For Cenozoic Earth we can substitute direct estimates of the important parameters into Equation~\eqref{eq:simple_milankovitch}.
Typical values for the time constant $\tau_R$ are of order $30 \, \mathrm{kyr}$ \citep{ricard1993polar}.
Estimates of the present day non-hydrostatic moment of inertia (due to mantle density anomalies, corresponding to $\Lambda_{31}I_0$ in the preceding scaling)
are in the neighborhood of $10^{-5} I_0$, while the hydrostatic moment of inertia (corresponding to $C-A$) is $3 \times 10^{-3} I_0$ \citep{chambat2001mean}.
A key question is whether convection is sufficiently chaotic to enter the large $\theta$ regime. 
\citet{richards1997explanation} have argued that the convective planform of Earth has been stable for the last
few hundred million years. On the other hand, we know that there have been large reorganizations of that planform 
during Earth history, so this recent geologic stability may not hold in general \citep{evans2003true}.
Our numerical simulations show that large values for $\theta$ are possible in a vigorously convecting mantle.
Allowing for such a large mismatch angle ($\theta = 45^\circ$) we may estimate the maximum polar wander rate
\begin{equation}
\max ( \dot{\Theta} ) = \frac{(\lambda_3-\lambda_1)}{(C-A)}\frac{1}{T_1} \sim 6^\circ / \mathrm{Myr},
\end{equation}
which corresponds to about 66 cm/yr 90$^\circ$ from the TPW axis.
This value is similar to the rates discussed by \citet{cambiotti2011new} for the past 100 Myr, though our rate is roughly a factor of four larger because we allow for the
possibility of a larger mismatch angle. Such large angles may be necessary to explain the range of rates suggested by some interpretations of paleomagnetic data \citep{mitchell2011sutton}.
The bulk viscosity of Earth's mantle is uncertain by up to a factor of ten \citep{mitrovica2004new},
which results in a corresponding uncertainty for the relaxation time $\tau_R$ and the maximum polar wander rate.
